\title{Desktop Awareness Project}
\author{
        Martin Renold \\
        Personal Research
        }
\date{\today}

% Title reasoning...
%Desktop Awareness Project (DAP)                      
%- who is aware? the users about the desktop? nope... 
%User Awareness Project (UAP)                         
%                                                     
%Desktop Sensor System (DSS) 
%Desktop Sensing System
%
% Alibi. - cool, but not googleable, and might be missinterpreted

\documentclass[]{article}

\begin{document}
\maketitle

\section{Introduction}

These days Desktop environments are mostly thought of as a dead
container for independent applications. Recent activity and the daily
workflow is knowlege reserved to humans. This project aims to monitor
the activity of the user. The data is made accessible to developers in
raw form or preprocessed for specific needs.

Examples are keyboard and mouse events, the name of the focused
application, recently started processes, or the amplitude recorded by
a microphone.

\subsection{Applications}
During the project, some sample applications are developed that use
this data. Here are some ideas:
\begin{itemize}
\item
Display lots of statistics, eg. the daily/weekly time spent using each
application, which applications are used together, the hourly uptime
statistic for the average day, etc.
\item
Provide information about user proximity for chat applications. Some
chat programs already have such functionality, but they will probably
not include microphone readings. It might even be possible to estimate
return times based on daily statistics in some cases. Address privacy
concerns.
\item
Allow the user to trigger commands on certain events, eg. when
entering/leaving an application. For example you could run scripts to
change the keyboard layout or mouse sensitivity for certain
programs only.
\item
Some email/chat/update programs provide baloon notifications. Hold
them back while the user is not present, and possibly wait until the
user stops typing before showing them.
\end{itemize}

\section{Data Sources}

The idea is to just throw as many different data sources as possible
at some central algorithm, and let it figure out whether they are
useful and how they are correlated. It is more important to have some
data from every source than to have it all processed properly.

\subsection{Keyboard and Mouse}

Keyboard statistics, maybe even some common keyboard shortcuts or
commands (careful about sensitive data...). Mouse motion in pixels
during the last 5 minutes, number of keys pressed during the lats
10seconds, minimum/maximum pause between keys/mouse.

\subsection{Microphone}
When available, the microphone will record many potentially useful
events. It can hear the user approaching and detect whether there is
anybody out there at all. It might even be able to distinguish the
sound of different doors and windows being opened or closed, and it
will most certainly also hear typing and clicking. It will also detect
chairs moving, vacuum cleaner, toilet flushing, cooking activity,
light switches being pressed, electrical disturbances, power supply
noise, music (possibly produced by the host PC), and many things you
would never have thought of.

Some signal processing is required to detect such events. As a first
approach, just calculate the average amplitude over some time, and its
derivative. Later, cepstral coefficients migth come handy.

\subsection{Video}
Video input, if available, produces a huge amount of data to process,
and a lot of literature to read about processing algorithms.  I
suggest to just average the brightness over four screen regions and
record their absolute and relative value and its derivative. This
should at least allow to detect when someone is passing by or moving.

There are tons of other features that could be added later, eg. the
noise level, color, sharpness, etc. and even the output of complete
object recognition algorithms (eg. viola jones face detection).

\subsection{System Statistics}
CPU load, harddisk activity, maybe CPU and Harddisk temperature.  New
processes created, process CPU and memory usage. System uptime, last
reboot, uptime last week, network connectivity, usual poweroff times,
etc.

\subsection{Window Manager}
Ask the Window manager about visible window, window with input focus,
when a window is created, etc.

\subsection{USB Bus}
Plugging of hardware devices. (eg. camera, usb sticks, etc)

\subsection{Audio Output}
Sounds being played by the PC. Beeps and music.

\subsection{Filenames}
A bit trickier: record whenever a file is saved. (To allow the user)

\subsection{Clipboard}
Copy/Paste of small text pieces.

\subsection{Application Specific Events}
Receiving a new email, sending one. URL/Title of the webpage being
surfed to. Person/Channel being chated with/in. A generic interface
could be provided where those applications can report such activity.

\section{Possible Applications}

\subsection{Application Usage Statistics}
Programs for that already exist on Windows. Display how many hours
each application was used. Can be useful eg. to monitor game
addiction.

Additionally, show an uptime graph with 24 hours on the x-axis and
number of days that the computer was up on the y-axis. Can do same
thing for individual applications.

\subsection{Chat Programs}
For chat programs, usually something like ``mark me idle after 5
minutes of inactivity'' is used. This could be improved by adding an
estimate about the return time, or eg. immediatly marking the user as
unavailable when the door to the room is closed, and mark the user as
available again when he enters the room (even without touching mouse
or keyboard). Privacy is a concern here, of course; just make sure the
user can choose exactly how much information he is giving away.

\subsection{User defined application starters}
If the recognized events can somehow be visualized to the user, the
use can be allowed to attach something to those events, eg. lock the
screen whenever the user stands up, or print a notification if there
was no vacum cleaner sound for more than two weeks, or shut down the
PC when the user left the building.

Extending the concept, the system might even detect repeated sequences
of actions and offer to automate the task. For example when the user
always opens the image in GIMP after taking a screenshot, the system
could ask whether it should do so automatically every time. Or focus
the email window whenever the user comes back.

\subsection{Background Tasks}
An API could be offered that finds out suitable start times for
starting background task like backups or indexing.

\subsection{Memory Refreshers}

[This might not belong into this project.]

Get memory refreshers about specific stuff, eg. if you view an email
of a person that you rarely read emails from, offer to view the
webpage where you last read about that same person. Same thing with
rare topics...

\section{Architecture}
hm...

\end{document}

